An Innovative Storage Stack Addressing the Requirements of Extreme Scale Platforms

### 2014-04-15 paper.tex
Adding description of ACG and HDF5 layer

\documentclass[conference]{IEEEtran}
\pdfpagewidth=8.5in
\pdfpageheight=11in
%\usepackage{subfig}
\usepackage{subfigure}
\usepackage[pdftex]{}
\usepackage{todonotes}
\usepackage{listings}
\lstset {% general command to set parameter(s)
         language=C,
     basicstyle=\footnotesize,               % print whole listing small
%     keywordstyle=\color{black}\bfseries, % underlined bold black keywords
%     identifierstyle =\color{black},  % nothing happens
%     commentstyle=\color{black}\emph, % white comments
     %stringstyle=\ttfamily,          % typewriter type for strings
%     stringstyle=\color{black},       % typewriter type for strings
         tabsize=4,
         showtabs=false,
     showstringspaces=false}
%can't figure this one out for particles bandwidth
%\usepackage[caption,label]{subfig}

@@

>>>>

\documentclass[conference]{IEEEtran}
\pdfpagewidth=8.5in
\pdfpageheight=11in
%\usepackage{subfig}
%\usepackage{subfigure}
\usepackage[pdftex]{}
\usepackage{graphicx}
%\usepackage{todonotes}
\usepackage{listings}
\lstset {% general command to set parameter(s)
         language=C,
     basicstyle=\footnotesize,               % print whole listing small
%     keywordstyle=\color{black}\bfseries, % underlined bold black keywords
%     identifierstyle =\color{black},  % nothing happens
%     commentstyle=\color{black}\emph, % white comments
     %stringstyle=\ttfamily,          % typewriter type for strings
%     stringstyle=\color{black},       % typewriter type for strings
         tabsize=4,
         showtabs=false,
     showstringspaces=false}
%can't figure this one out for particles bandwidth
%\usepackage[caption,label]{subfig}

@@

%%%%%%%%%%%%%%%%%%%%%%%%% <<<<<<<<<<<<<<<<

The ``I/O nodes'' handle 
requests forwarded by the scientific applications, integrate a tier of
solid-state devices to absorb the burst of random operations, and
organize/re-format the data so that transfers from the staging area to
the traditional parallel file system can be done more efficiently. It
also has the capacity of executing analysis on the fresh data that has
just been generated by simulation applications running at the compute
nodes. As the data arrives, re-organization and data preparation can be
applied in order to anticipate the execution of analytical tasks
{[}3{]}.

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.30]{images/exa-arch.png}
\caption{Exascale Architecture. \label{exa-arch}}
\end{figure}

A common cluster configuration for this type of deployment is shown in
Figure \ref{exa-arch}. The designated I/O nodes (IONs) are connected to
the compute nodes (CNs) through the same fast fabric (eg. Infiniband),
while the connection to the external storage cluster is through a
secondary slower channel (eg. 10Gb ethernet). I/O staging thus
introduces a new layer in the application-to-storage stack, as shown in
Figure \ref{exa-stack} (left).

In current proposals, the stack is managed by middleware that seats
between the application and the parallel file system {[}2,4{]}. Since
most of the applications assume a POSIX interface to storage {[}5{]},
existing middleware either transparently handle the I/O operations,
appearing as regular POSIX calls to applications; or modify the I/O API
as little as possible in order to minimize the impact on production
codes.

As we move towards the exascale goal, this way of interfacing with
storage becomes more of an obstacle. Many features provided by the stack
are hidden by the middleware/POSIX layers. If these were visible to the
applications, and if the applications were able to control how the
underlying layers behave, they would benefit greatly since the domain
knowledge that is available at the application level could be used to
execute storage operations efficiently. One more implication of the
25-year-old POSIX API is in cumbersome middleware, since metadata has to
be part of the same uni-dimensional byte array being stored.

The Fast Forward Storage I/O project {[}6{]} is aimed at merging the
features of existing middleware into a next-generation parallel file
system stack. This new storage API termed the I/O dispatcher (IOD) is
object-based, transactional, asynchronous and I/O-staging-aware (Figure
\ref{exa-stack} (right)).

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.30]{images/exa-stack.png}
\caption{Top-to-bottom view of the component stack. \label{exa-stack}}
\end{figure}

The prototype stack, from a top-down point of view, has the following
components:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Arbitrarily Connected Graphs (ACG).
\item
  HDF5 extensions.
\item
  I/O Dispatcher (IOD).
\item
  Distributed Application Object Storage (DAOS).
\item
  Versioning Object Storage Device (VOSD).
\end{enumerate}

Applications running on the compute nodes will be written in Python
scripts or in C++ using the HDF5 APIs. These applications will execute
analysis on graph-based data (ACG). The point is to demonstrate both the
HPC and BigData use cases of the Exascale architecture. The data
structures will be stored in HDF5, for which new API extensions will be
implemented. These new API calls will expose the transactional,
asynchronous and function-forwarding semantics of the underlying stack.

As mentioned before, the stack provides \emph{non-POSIX, object-based,
transactional and asynchronous active-storage}, meaning that POSIX is
supplanted by new object interfaces that reach up to the HDF5 layer
(Application I/O in the stack diagram); transactional semantics are
present, from HDF down to the VOSD layer (Storage layer); clients don't
need to wait for any blocking operation; and analysis can be shipped and
executed on I/O or DAOS nodes.

We give a high-level description of each of layer.

\subsection{ACG}\label{acg}

This will exemplify how applications make use of the Exascale stack. For
FF, support for
GraphLab 
and GraphBuilder will be prototyped, which are graph-processing
frameworks. GraphBuilder is a set of MapReduce tasks that extract,
normalize, partition and serialize a graph out of unstructured data, and
writes graph-specific formats into HDFS. These files are later consumed
by GraphLab, a vertex-centric, asynchronous execution engine that runs
directly on top of HDFS (i.e. non-MapReduce). The following illustrates
the architecture of both frameworks:

\begin{figure}[htbp]
\centering
\includegraphics{images/graphlab-and-graphbuilder.png}
\caption{Figure 3. GraphLab and GraphBuilder stacks (taken from
{[}4{]}).}
\end{figure}

In order to make both work on top of the exascale stack, both have to be
modified. After these modifications are implemented, GraphBuilder will
be able to write the partitioned graph in (the newly proposed) HDF5
files which will thus be stored in the IOD nodes (or IONs) in a
parallel-optimized way. On the GraphLab side, HDF5-awareness will allow
the library to perform at high speeds by benefiting from the new
features (see next section). In general both frameworks will be modified
so that calls to HDFS-based formats are replaced by the proposed HDF5
ones. This is referred to as the HDF Adaptation Layer or HAL and will
provide, from the GraphBuilder/GraphLab point of view {[}5{]}:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  capability for storing the newly proposed HDF5 format
\item
  association of network information to vertices/edges
\item
  shipping computation to the IONs
\item
  asynchronous vertex updates
\item
  efficient data sharing ammong CNs
\item
  computation over versioned datasets
\end{itemize}

\subsection{HDF5 extensions}\label{hdf5-extensions}

The extensions done to HDF5 allows an application to take full advantage
of the new exascale features. The additions comprise {[}6{]} (Figure 4):

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Object-storage API based on HDF5 to support high-level data models.
  This exposes asynchronous, transactional semantics to the application,
  as well as end-to-end data integrity. It will also allows the usage of
  pointer data types, passing of hints down to the storage and support
  for asynchronous index building, maintenance and querying.
\item
  Virtual Object Layer (VOL) plugin that translates HDF5 API requests
  from applications to IOD API calls.
\item
  Function shipping from CN to IONs. This provides the application
  developer with the capacity of sending computation down to the IONs
  and get back results.
\item
  Analysis Shipping from CN to IONs or DAOS nodes. This is similar to 3
  but instead of returning the result over the network, it gets stored
  on the nodes and pointers to it are returned.
\end{enumerate}

In terms of the CN-ION communication model, a client/server architecture
is implemented {[}8{]}: every ION runs an IOFSL (I/O Function Shipping
Layer) server {[}9{]}; the IOFSL client is integrated into the HDF5
library which runs on each CN. A client can forward requests to any
number of IONs. Every I/O operation issued by HDF5 is asynchronously
shipped to the IOFSL server and asynchronously executed.

HDF5 {[}7{]} is a versatile data model containing complex data objects
and metadata. Its information set is a collection of datasets, groups,
datatypes and metadata objects. The data model defines mechanisms for
creating associations between various information items. The main
components of HDF5 are described below.

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.30]{images/hdf5.png}
\caption{Sample file illustrating the hierarchical structure of HDF5
files. \label{hdf5-sample}}
\end{figure}

\begin{itemize}
\item
  \textbf{File}: In the HDF5 data model the container of an HDF5 infoset
  is represented by a file. It is a collection of objects that also
  explains the relationship between them. Every file begins with a root
  group ''/'', which serves as the ''starting-point'' in the object
  hierarchy.
\item
  \textbf{Dataset}: HDF5 datasets are objects that represent actual data
  or content. Datasets are arrays which can have multiple dimensions. A
  dataset is characterized by a dataspace and a datatype. The dataspace
  captures the rank (number of dimensions), and the current and maximum
  extent in each dimension. The datatype describes the type of its data
  elements.
\item
  \textbf{Group}: A group is an explicit association between HDF5
  objects. It is synonymous with directories in a file system. A group
  could contain multiple other groups, datasets or datatypes within it.
\item
  \textbf{Attribute}: Attributes are used for annotating datasets,
  groups, and datatype objects. They are datasets themselves, and are
  attached to existing objects they annotate.
\end{itemize}

For example, as shown in Figure \ref{hdf5-sample}, the file
Sample.h5 contains the root group which itself contains a
group G1 and two datasets, D1 and D2. Group
G1 contains a dataset D3. Attribute A1 is
linked to dataset D1. The objects and the relationships between
them can be represented as a B-tree, which is used internally by HDF5 to
index its objects.

The Fast Forward project is working on the adaptation of the HDF5
libraries in order to expose the storage stack to applications.

\subsection{Virtual Object Layer}\label{virtual-object-layer}

The Virtual Object Layer is an abstraction mechanism internal to the
HDF5 library {[}\textbf{???}{]}. As shown in Figure \ref{vol-arch} it is
implemented just below the public API. The VOL exports an interface that
allows writing plugins for HDF5, thereby enabling developers to store
objects in a format different from the default HDF5 file format. Plugin
writers provide an implementation for a set of functions that access
data on disk or on a parallel file system.

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.20]{images/vol-arch.png}
\caption{Architectural view of the VOL abstraction mechanism.
\label{vol-arch}}
\end{figure}

\subsection{IOD VOL plugin}\label{iod-vol-plugin}

The IOD VOL plugin serves as the bridge between HDF5 and the IOD API. It
is also shown in Figure \ref{vol-arch}. The application calls the HDF5
library while running on the system's compute nodes. Using the VOL
architecture, the IOD VOL plugin uses a function shipper (RPC library)
to forward the VOL calls to a server component running on the I/O nodes
(IONs). At that point, the VOL calls are translated into I/O Dispatcher
(IOD) API calls and executed at the IONs.

The mapping between HDF5 objects and IOD objects is the following:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  file = container
\item
  dataset = array
\item
  group = key-value store
\item
  attribute = key-value store
\end{itemize}

\subsection{Fast Forward's I/O Dispatcher
API}\label{fast-forwards-io-dispatcher-api}

The Fast Forward I/O and Storage project is aimed at integrating many of
the services existing currently in middleware libraries in a
next-generation storage stack, exposed through a non-POSIX,
transactional, asynchronous, object-based API.

From a high-level point of view, the Fast Forward's I/O Dispatcher (IOD)
interface allows the user to:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  transactions
\item
  asynchrony
\item
  objects
\item
  layout
\item
  placement
\item
  formats
\end{itemize}

Type of objects:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  container
\item
  key-value store
\item
  blobs
\item
  multi-dimensional arrays
\end{itemize}

As mentioned previously, IOD allows to store distinct types of objects.
From here on we will limit our discussion to multidimensional arrays
(HDF5 datasets) and leave the treatment of other object types as future
work. A multidimensional array can be sharded among the distinct I/O
nodes in the staging area. The IOD API supports the following
strategies:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \textbf{contiguous}. fixed chunking, distributed in a round-robin
  fashion accross the I/O nodes.
\item
  \textbf{chunked}. same as above but with irregular (sparse) chunking.
\item
  \textbf{user-defined}. either contiguous or chunked but user specifies
  where to place each individual shard.
\end{itemize}

It is possible to request the transformation of an object's physical
layout to other formats, having multiple copies of the same objects in
multiple formats if desired. Also, the user can pre-fetch objects from
the storage cluster into the I/O nodes or read them directly from the
storage cluster. At the semantic level (HDF5), indexes can be created
for datasets, which results in being able to read through an index
instead of directly from the base array.

{[}sharding. \label{sharding}{]}{[}sharding{]}

All these distinct alternatives result in having many different ways of
executing the same analysis task. In the future, we plan to extend our
design to incorporate an optimization phase that will prepare a task by
taking into account the distinct possible ways of executing and select
the most promising one. In our subsequent discussions, we consider only
data-movement optimization, i.e.~sending the analysis code as close as
possible to the data. In practice, this means we focus on identifying
sharding of datasets and execute code accordingly over the appropriate
shards.

%%%%%%%%%%%%%%%%%%%%%%%%% >>>>>>>>>>>>>>
